Practical Optional Types for Clojure

### 2015-10-13 overview.tex
convert to ESOP format
fix esop formatting

\begin{minted}{clj}
(ann  greet [Str -> Str])
(defn greet [n]
  (str "Hello, " n "!"))
(greet "Grace") ; :- Str
;=> "Hello, Grace!"
\end{minted}
%
Strings are accepted, but providing \clj{nil} (Clojure's name for Java's \java{null})
is a static type error---\clj{nil} is not a string in Typed Clojure.
%
\begin{minted}{clj}
(greet nil) ; Type Error:
            ;   Expected Str, given nil
\end{minted}

@@

\paragraph{Flow analysis} Typed Clojure uses occurrence typing~\cite{TF10} to
model type-based control flow.
In \clj{greetings}, a \clj{when} expression ensures \clj{repeat}
is never passed a \clj{nil} argument.
%
\begin{minted}{clj}
(ann  greetings [Str (U nil Int) -> Str])
(defn greetings [n i]
  (str "Hello, "
       (when i  ; when i is a non-nil integer
         (apply str (repeat i "hello, ")))
       n "!"))
(greetings "Donald" 2)  ; :- Str
;=> "Hello, hello, hello, Donald!"
(greetings "Grace" nil) ; :- Str
;=> "Hello, Grace!"
\end{minted}
%
Removing the \clj{when} expression is a static type error---\clj{repeat} 
cannot be passed \clj{nil}.
%
\begin{minted}{clj}
(ann  greetings-bad [Str (U nil Int) -> Str])
(defn greetings-bad [n i]
  (str "Hello, "
       (apply str 
         (repeat 
           i ; Type Error:
             ;   Expected Int, given (U nil Int).
           "hello, "))
       n "!"))
\end{minted}

%\subsection{Type System Basics}
%
%\citet{TF10}
%presented Typed Racket with occurrence typing,
%a technique for deriving type information from conditional control flow.
%They introduced the concept of occurrence typing 
%with the following example.
%
%\inputminted[firstline=3]{racket}{code/tr/example1.rkt}
%
%This function takes a value that is either \emph{\#f} % mintinline really hates #
%or a number, represented by an \emph{untagged} union type.
%The `then' branch has an implicit invariant
%that \rkt{x} is a number, which is automatically inferred with occurrence typing
%and type checked without further annotations.
%
%We chose to build on the ideas and implementation
%of Typed Racket to implement a type system targeting Clojure for several reasons.
%Initially, the similarities between Racket and Clojure drew us to
%investigate the effectiveness of repurposing occurrence typing
%for a Clojure type system---both languages share a Lisp heritage,
%similar standard functions 
%(for instance \clj{map}
%in both languages is variable-arity)
%and idioms.
%While Typed Racket is gradually typed and has sophisticated
%dynamic semantics for cross-language interaction, we 
%chose to first implement
%the static semantics
%with the hope to extend Typed Clojure to be gradually typed at a future date.
%Finally,
%Typed Racket's combination of bidirectional checking
%and occurrence typing presents a successful model for 
%type checking dynamically typed programs without compromising
%soundness, which is appealing over success typing~\cite{Lindahl:2006:PTI}
%which cannot prove strong properties about programs
%and soft typing~\cite{CF91}
%which has proved too complicated in practice.
%
%Here is the above program in Typed Clojure.
%\begin{exmp}
%\inputminted[firstline=5]{clojure}{code/demo/src/demo/eg1.clj}
%\label{example:conditionalflow}
%\end{exmp}
%
%The \clj{fn} macro (provided by core.typed) supports optional annotations by 
%adding
%\clj{:-} and a type after a parameter
%position
%or binding vector 
%to annotate parameter types
%and return types respectively.
%\clj{number?} is
%a Java \java{instanceof} test of \clj{java.lang.Number}.
%As in Typed Racket, \clj{U} creates an \emph{untagged union} type, which can take
%any number of types.
%
%Typed Clojure can already check all of the examples in~\citet{TF10}---the 
%rest of this section describes the extensions necessary
%to check Clojure code.

@@

\begin{minted}{clj}
(fn [f]           ; Type Error:                    
  (.getParent f)) ;   Unresolved interop: getParent

@@

The conditional guards from dereferencing \clj{nil}, and---as before---removing 
it is a static type error, as typed code could possibly dereference \clj{nil}.
\begin{minted}{clj}
(defn parent-bad-in [f :- (U nil File)]
  (.getParent f)) ; Type Error:
                  ;   Cannot call instance method 
                  ;   java.io.File/getParent on type 
                  ;   (U nil File).
\end{minted}
Since Java-level types do not provide information about nullability,
by default Typed Clojure conservatively assumes that Java method and
constructor arguments, such as \clj{"a/b"} provided to the \clj{File} constructor in
\egref{example:getparent-direct-constructor},
 must be \emph{non-nullable}, but this can be
configured for  particular calls as needed.
The target of method invocations is always non-nullable.

Any Java method returning a reference can also return \java{null} ---
Typed Clojure rejects programs that assume otherwise.
%
\begin{minted}{clj}
(defn parent-bad-out [f :- File] :- Str
  (.getParent f)) ; Type Error:
                  ;   Expected Str, given (U nil Str).
\end{minted}

@@

Hash-maps are accessed with the \clj{get} function:
\begin{exmp}
\begin{minted}{clojure}
(def breakfast 
  {:en "waffles" :fr "croissants"})

(get breakfast :en)    ; :- Str
;=> "waffles"
\end{minted}
\label{example:breakfastcomplete}
\end{exmp}
Additionally, keywords are functions that look themselves up in a
map. Missing keys produce \clj{nil}.
\begin{minted}{clojure}
(:fr breakfast)        ; :- Str
;=> "croissants"
(:bocce breakfast) ; :- nil
;=> nil
\end{minted}

In Typed Clojure, \emph{HMap types} describe the most common usages of
keyword-keyed maps.
\begin{minted}{clojure}
breakfast ; :- (HMap :mandatory {:en Str :fr Str}
          ;          :complete? true)
\end{minted}
The inferred type for \clj{breakfast} holds two kinds of information---the known entries--\clj{:en}
and \clj{:fr}---and their types, which are  \clj{:mandatory}, and that 
no other key is present, since \clj{:complete?} is \clj{true}.

HMap types default to  partial specification---\clj{:complete?} defaults to \clj{false}. The HMap shorthand 
\clj{'{:en Str :fr Str}}
omits
information about absent keys, only providing information on
\clj{:mandatory} keys.
%
\begin{exmp}
\begin{minted}{clojure}
(ann lunch '{:en Str :fr Str})
(def lunch {:en "muffin" :fr "baguette"})
(:en lunch)    ; :- Str
;=> "muffin"
(:fr lunch)    ; :- Str
;=> "baguette"
; Unknown lookups are now less accurate
(:bocce lunch) ; :- Any
;=> nil
\end{minted}
\label{example:lunchpartial}
\end{exmp}

@@

\begin{exmp}
\begin{minted}{clojure}
(defalias RawKeyPair
  "Unencrypted keypair -- extra keys disallowed"
  (HMap :mandatory {:public-key RawKey,
                    :private-key RawKey},
        :complete? true))

(defalias EncKeyPair
  "Encrypted keypair -- extra keys disallowed"
  (HMap :mandatory {:public-key RawKey,
                    :enc-private-key EncKey},
        :complete? true))

(ann enc-keypair [RawKeyPair -> EncKeyPair])
(defn enc-keypair "Encrypt an unencrypted keypair"
  [{pk :private-key :as kp}] ; original map is kp
  (assoc 
    ; remove unencrypted private key
    (dissoc kp :private-key)
    ; add encrypted private key
    :enc-private-key (encrypt pk)))
\end{minted}
\label{example:circleci}
\end{exmp}

%\inputminted[firstline=10,lastline=22]{clojure}{code/demo/src/demo/key.clj}
If we forget to remove the unencrypted private key,
a type error is given, because \clj{EncKeyPair} is fully specified.
%
\begin{minted}{clojure}
(ann enc-keypair-bad [RawKeyPair -> EncKeyPair])
(defn enc-keypair-bad
  [{pk :private-key :as kp}]
  (assoc kp :enc-private-key (encrypt pk)))
  ; Type Error:
  ;   Expected EncKeyPair, given 
  ;   (HMap :mandatory {:enc-private-key EncKey
  ;                     :private-key RawKey
  ;                     :public-key RawKey}
  ;         :complete? true)
\end{minted}
The extra \clj{:private-key} entry does not match \clj{EncKeyPair}, so a type error
is raised.

%\clj{enc-keypair} takes an unencrypted keypair and returns an encrypted keypair by
%dissociating the raw \clj{:private-key} entry with \clj{dissoc}
%and associating an encrypted private key
%as \clj{:enc-private-key} on an immutable map with \clj{assoc}.
%The expression \clj{(:private-key kp)} shows that keywords are also 
%functions that look themselves up in a map returning the associated value or \nil{} if the key is missing.
%Since \clj{EncKeyPair} is \clj{:complete?}, Typed Clojure enforces the return type
%does not contain an entry \clj{:private-key}, and would complain if the \clj{dissoc}
%operation forgot to remove it.

%\egref{example:absentkeys}
%is like \egref{example:circleci}
%except the \clj{:absent-keys} HMap option is used
%instead of \clj{:complete?},
%which takes a \emph{set literal} of keywords that do not appear in the map, written 
%with \emph{\#}-prefixed braces.
%The syntax \clj{(fn [{pkey :private-key, :as kp}] ...)}
%aliases \clj{kp} to the first argument and \clj{pkey} to \clj{(:private-key m)}
%in the function body.
%
%\begin{exmp}
%\inputminted[firstline=10,lastline=21]{clojure}{code/demo/src/demo/key2.clj}
%\label{example:absentkeys}
%\end{exmp}
%
%Since this example enforces that \clj{:private-key} must not appear
%in a \clj{EncKeyPair}
%Typed Clojure would still complain if we forgot to \clj{dissoc} \clj{:private-key}
%from the return value.
%Now, however we could stash the raw private key in another entry
%like \clj{:secret-key} which is not mentioned by the partial HMap \clj{EncKeyPair}
%without Typed Clojure noticing.

@@

\begin{minted}{clojure}
(defalias Order   ; define type abbreviation
  "A meal order, tracking dessert quantities."
  (U '{:Meal ':lunch ; keyword singleton type
       :desserts Int}
     '{:Meal ':dinner :desserts Int}
     '{:Meal ':combo :meal1 Order :meal2 Order}))
\end{minted}

@@

\begin{exmp}
\begin{minted}{clojure}
(ann desserts [Order -> Int])
(defmulti desserts "Total desserts per order."
          :Meal)  ; dispatch on :Meal entry
; destructuring reads backwards, d is :desserts entry
(defmethod desserts :lunch [{d :desserts}] d)
(defmethod desserts :dinner [{d :desserts}] d)
(defmethod desserts :combo [{m1 :meal1 m2 :meal2}]
  (+ (desserts m1) (desserts m2)))

(desserts {:Meal :combo 
           :meal1 {:Meal :lunch :desserts 1}
           :meal2 {:Meal :dinner :desserts 2}})
;=> 3
\end{minted}
\label{example:desserts-on-meal}
\end{exmp}

@@

\begin{exmp}
\begin{minted}{clojure}
(ann desserts' [Order -> Int])
(defmulti desserts' 
          (fn [o :- Order] (class (:desserts o))))
(defmethod desserts' Long [{d :desserts :as o}] 
  ; o :- (U '{:Meal ':lunch :desserts Int}
  ;         '{:Meal ':dinner :desserts Int}
  ;         '{:Meal ':combo :desserts Int
  ;           :meal1 Order :meal2 Order})
  d)
(defmethod desserts' nil [{m1 :meal1 m2 :meal2 :as o}]
  ; o :- '{:Meal ':combo :meal1 Order :meal2 Order}
  (+ (desserts' m1) (desserts' m2)))
\end{minted}
\label{example:desserts-on-class}
\end{exmp}
%
%(desserts' {:Meal :combo 
%            :meal1 {:Meal :lunch :desserts 1}
%            :meal2 {:Meal :dinner :desserts 2}})
%;=> 3

@@

In the \clj{nil} method,
\clj{(isa? (class (:desserts o)) nil)}
must be true---which implies \clj{(class (:desserts o))} is \clj{nil}.
%
Since lookups on missing keys return \clj{nil}, either
\begin{itemize}
  \item \clj{o} has a \clj{:desserts} entry to \clj{nil}, like \clj{{:desserts nil}}, or
  \item \clj{o} is missing a \clj{:desserts} entry, like \clj{{}}.
\end{itemize}
Equivalently, we learn \clj{o} is at least of type
\begin{minted}{clojure}
(U '{:desserts nil}
   ; :absent-keys, a set of known absent entries
   (HMap :absent-keys #{:desserts}))
\end{minted}
This eliminates non-\clj{:combo} meals
since their \clj{'{:desserts Int}} type does not agree
with this new information (because \clj{:desserts}
is neither mapped to \clj{nil} or absent).

>>>>

\begin{minted}{clj}
(ann  greet [Str -> Str])
(defn greet [n] (str "Hello, " n "!"))
(greet "Grace") ; :- Str
;=> "Hello, Grace!"
\end{minted}
%
Strings are accepted, but providing \clj{nil} (Clojure's name for Java's \java{null})
is a static type error---\clj{nil} is not a string in Typed Clojure.
%
\begin{minted}{clj}
(greet nil) ; Type Error: Expected Str, given nil
\end{minted}

@@

\paragraph{Flow analysis} Typed Clojure uses occurrence typing~\cite{TF10} to
model type-based control flow.
In \clj{greetings}, a \clj{when} expression ensures \clj{repeat}
is never passed a \clj{nil} argument.
%
\begin{minted}{clj}
(ann  greetings [Str (U nil Int) -> Str])
(defn greetings [n i]
  (str "Hello, " (when i  ; when i is a non-nil integer
                   (apply str (repeat i "hello, ")))
       n "!"))
(greetings "Donald" 2)  ; :- Str
;=> "Hello, hello, hello, Donald!"
(greetings "Grace" nil) ; :- Str
;=> "Hello, Grace!"
\end{minted}
%
Removing the \clj{when} expression is a static type error---\clj{repeat} 
cannot be passed \clj{nil}.
%
\begin{minted}{clj}
(ann  greetings-bad [Str (U nil Int) -> Str])
(defn greetings-bad [n i]
  (str "Hello, " (apply str 
                   (repeat i ; Type Error: Expected Int, given (U nil Int).
                           "hello, "))
       n "!"))
\end{minted}

%\subsection{Type System Basics}
%
%\cite{TF10}
%presented Typed Racket with occurrence typing,
%a technique for deriving type information from conditional control flow.
%They introduced the concept of occurrence typing 
%with the following example.
%
%\inputminted[firstline=3]{racket}{code/tr/example1.rkt}
%
%This function takes a value that is either \emph{\#f} % mintinline really hates #
%or a number, represented by an \emph{untagged} union type.
%The `then' branch has an implicit invariant
%that \rkt{x} is a number, which is automatically inferred with occurrence typing
%and type checked without further annotations.
%
%We chose to build on the ideas and implementation
%of Typed Racket to implement a type system targeting Clojure for several reasons.
%Initially, the similarities between Racket and Clojure drew us to
%investigate the effectiveness of repurposing occurrence typing
%for a Clojure type system---both languages share a Lisp heritage,
%similar standard functions 
%(for instance \clj{map}
%in both languages is variable-arity)
%and idioms.
%While Typed Racket is gradually typed and has sophisticated
%dynamic semantics for cross-language interaction, we 
%chose to first implement
%the static semantics
%with the hope to extend Typed Clojure to be gradually typed at a future date.
%Finally,
%Typed Racket's combination of bidirectional checking
%and occurrence typing presents a successful model for 
%type checking dynamically typed programs without compromising
%soundness, which is appealing over success typing~\cite{Lindahl:2006:PTI}
%which cannot prove strong properties about programs
%and soft typing~\cite{CF91}
%which has proved too complicated in practice.
%
%Here is the above program in Typed Clojure.
%\begin{exmp}
%\inputminted[firstline=5]{clojure}{code/demo/src/demo/eg1.clj}
%\label{example:conditionalflow}
%\end{exmp}
%
%The \clj{fn} macro (provided by core.typed) supports optional annotations by 
%adding
%\clj{:-} and a type after a parameter
%position
%or binding vector 
%to annotate parameter types
%and return types respectively.
%\clj{number?} is
%a Java \java{instanceof} test of \clj{java.lang.Number}.
%As in Typed Racket, \clj{U} creates an \emph{untagged union} type, which can take
%any number of types.
%
%Typed Clojure can already check all of the examples in~\cite{TF10}---the 
%rest of this section describes the extensions necessary
%to check Clojure code.

@@

\begin{minted}{clj}
(fn [f] (.getParent f)) ; Type Error: Unresolved interop: getParent

@@

The conditional guards from dereferencing \clj{nil}, and---as before---removing 
it is a static type error, as typed code could possibly dereference \clj{nil}.
\begin{minted}{clj}
(defn parent-bad-in [f :- (U nil File)]
  (.getParent f)) ; Type Error: Cannot call instance method 
                  ; java.io.File/getParent on type (U nil File).
\end{minted}
Since Java-level types do not provide information about nullability,
by default Typed Clojure conservatively assumes that Java method and
constructor arguments, such as \clj{"a/b"} provided to the \clj{File} constructor in
\egref{example:getparent-direct-constructor},
 must be \emph{non-nullable}, but this can be
configured for  particular calls as needed.
The target of method invocations is always non-nullable.

Any Java method returning a reference can also return \java{null} ---
Typed Clojure rejects programs that assume otherwise.
%
\begin{minted}{clj}
(defn parent-bad-out [f :- File] :- Str
  (.getParent f)) ; Type Error: Expected Str, given (U nil Str).
\end{minted}

@@

Hash-maps are accessed with the \clj{get} function:
\begin{exmp}
\begin{minted}{clojure}
(def breakfast {:en "waffles" :fr "croissants"})
(get breakfast :en)    ; :- Str
;=> "waffles"
\end{minted}
\label{example:breakfastcomplete}
\end{exmp}
Additionally, keywords are functions that look themselves up in a
map. Missing keys produce \clj{nil}.
\begin{minted}{clojure}
(:fr breakfast)        ; :- Str
;=> "croissants"
(:bocce breakfast) ; :- nil
;=> nil
\end{minted}

In Typed Clojure, \emph{HMap types} describe the most common usages of
keyword-keyed maps.
\begin{minted}{clojure}
breakfast ; :- (HMap :mandatory {:en Str, :fr Str}, :complete? true)
\end{minted}
The inferred type for \clj{breakfast} holds two kinds of information---the known entries--\clj{:en}
and \clj{:fr}---and their types, which are  \clj{:mandatory}, and that 
no other key is present, since \clj{:complete?} is \clj{true}.

HMap types default to  partial specification---\clj{:complete?} defaults to \clj{false}. The HMap shorthand 
\clj{'{:en Str :fr Str}}
omits
information about absent keys, only providing information on
\clj{:mandatory} keys.
%
\begin{exmp}
\begin{minted}{clojure}
(ann lunch '{:en Str :fr Str})
(def lunch {:en "muffin" :fr "baguette"})
(:en lunch)    ; :- Str
;=> "muffin"
(:fr lunch)    ; :- Str
;=> "baguette"
(:bocce lunch) ; :- Any     ; Unknown lookups are now less accurate
;=> nil
\end{minted}
\label{example:lunchpartial}
\end{exmp}

@@

\begin{exmp}
\begin{minted}{clojure}
(defalias RawKeyPair ; extra keys disallowed
  (HMap :mandatory {:pub RawKey, :priv RawKey}, :complete? true))

(defalias EncKeyPair ; extra keys disallowed
  (HMap :mandatory {:pub RawKey, :enc-priv EncKey}, :complete? true))

(ann enc-keypair [RawKeyPair -> EncKeyPair])
(defn enc-keypair [{pk :priv :as kp}] ; original map is kp
  (assoc (dissoc kp :priv)       ; remove unencrypted private key
         :enc-priv (encrypt pk))) ; add encrypted private key
\end{minted}
\label{example:circleci}
\end{exmp}

%\inputminted[firstline=10,lastline=22]{clojure}{code/demo/src/demo/key.clj}
If we forget to remove the unencrypted private key,
a type error is given, because \clj{EncKeyPair} is fully specified.
%
\begin{minted}{clojure}
(ann enc-keypair-bad [RawKeyPair -> EncKeyPair])
(defn enc-keypair-bad [{pk :priv :as kp}]
  (assoc kp :enc-priv (encrypt pk)))
  ; Type Error: Expected EncKeyPair, given 
  ;   (HMap :mandatory {:enc-priv EncKey, :priv RawKey, :pub RawKey}
  ;         :complete? true)
\end{minted}
The extra \clj{:priv} entry does not match \clj{EncKeyPair}, so a type error
is raised.

%\clj{enc-keypair} takes an unencrypted keypair and returns an encrypted keypair by
%dissociating the raw \clj{:priv} entry with \clj{dissoc}
%and associating an encrypted private key
%as \clj{:enc-priv} on an immutable map with \clj{assoc}.
%The expression \clj{(:priv kp)} shows that keywords are also 
%functions that look themselves up in a map returning the associated value or \nil{} if the key is missing.
%Since \clj{EncKeyPair} is \clj{:complete?}, Typed Clojure enforces the return type
%does not contain an entry \clj{:priv}, and would complain if the \clj{dissoc}
%operation forgot to remove it.

%\egref{example:absentkeys}
%is like \egref{example:circleci}
%except the \clj{:absent-keys} HMap option is used
%instead of \clj{:complete?},
%which takes a \emph{set literal} of keywords that do not appear in the map, written 
%with \emph{\#}-prefixed braces.
%The syntax \clj{(fn [{pkey :priv, :as kp}] ...)}
%aliases \clj{kp} to the first argument and \clj{pkey} to \clj{(:priv m)}
%in the function body.
%
%\begin{exmp}
%\inputminted[firstline=10,lastline=21]{clojure}{code/demo/src/demo/key2.clj}
%\label{example:absentkeys}
%\end{exmp}
%
%Since this example enforces that \clj{:priv} must not appear
%in a \clj{EncKeyPair}
%Typed Clojure would still complain if we forgot to \clj{dissoc} \clj{:priv}
%from the return value.
%Now, however we could stash the raw private key in another entry
%like \clj{:secret-key} which is not mentioned by the partial HMap \clj{EncKeyPair}
%without Typed Clojure noticing.

@@

%FIXME define defalias above and keyword singletons

\begin{minted}{clojure}
(defalias Order "A meal order, tracking dessert quantities."
  (U '{:Meal ':lunch, :desserts Int} '{:Meal ':dinner :desserts Int}
     '{:Meal ':combo :meal1 Order :meal2 Order}))
\end{minted}

@@

\begin{exmp}
\begin{minted}{clojure}
(ann desserts [Order -> Int])
(defmulti desserts :Meal)  ; dispatch on :Meal entry
(defmethod desserts :lunch [{d :desserts}] d) ; d is :desserts entry
(defmethod desserts :dinner [{d :desserts}] d)
(defmethod desserts :combo [{m1 :meal1 m2 :meal2}] 
  (+ (desserts m1) (desserts m2)))

(desserts {:Meal :combo, :meal1 {:Meal :lunch :desserts 1}, 
           :meal2 {:Meal :dinner :desserts 2}})
;=> 3
\end{minted}
\label{example:desserts-on-meal}
\end{exmp}

@@

\begin{exmp}
\begin{minted}{clojure}
(ann desserts' [Order -> Int])
(defmulti desserts' (fn [o :- Order] (class (:desserts o))))
(defmethod desserts' Long [{d :desserts :as o}] 
  ; o :- (U '{:Meal ':lunch, :desserts Int} '{:Meal ':dinner, :desserts Int}
  ;         '{:Meal ':combo :desserts Int, :meal1 Order, :meal2 Order})
  d)
(defmethod desserts' nil [{m1 :meal1 m2 :meal2 :as o}]
  ; o :- '{:Meal ':combo, :meal1 Order, :meal2 Order}
  (+ (desserts' m1) (desserts' m2)))
\end{minted}
\label{example:desserts-on-class}
\end{exmp}
%
%(desserts' {:Meal :combo 
%            :meal1 {:Meal :lunch :desserts 1}
%            :meal2 {:Meal :dinner :desserts 2}})
%;=> 3

@@

In the \clj{nil} method,
\clj{(isa? (class (:desserts o)) nil)}
must be true---which implies \clj{(class (:desserts o))} is \clj{nil}.
%
Since lookups on missing keys return \clj{nil}, either
\begin{itemize}
  \item \clj{o} has a \clj{:desserts} entry to \clj{nil}, like \clj{{:desserts nil}}, or
  \item \clj{o} is missing a \clj{:desserts} entry, like \clj{{}}.
\end{itemize}
Equivalently, we learn \clj{o} is of type
% Note: mintedinline doesn't work with hash characters #
\begin{minted}{clojure}
; :absent-keys, a set of known absent entries
(U '{:desserts nil} (HMap :absent-keys #{:desserts}))
\end{minted}
This eliminates non-\clj{:combo} meals
since their \clj{'{:desserts Int}} type does not agree
with this new information (because \clj{:desserts}
is neither mapped to \clj{nil} or absent).

### 2015-10-13 related-work.tex
fix esop formatting

% Cite a few of the early papers here.
%http://www.cs.washington.edu/research/projects/cecil/www/pubs/
\paragraph{Multimethods} 
\cite{MS02} and collaborators present a sequence of
systems~\cite{Chambers:1992:OMC, Chambers:1994:TMM, MS02} with statically-typed multimethods
and modular type checking.  In contrast to Typed Clojure, in these
system methods declare the types of arguments that they expect which
corresponds to exclusively using \clj{class} as the dispatch function
in Typed Clojure. However, Typed Clojure does not attempt to rule out
failed dispatches at runtime.

>>>>

% Cite a few of the early papers here.
%http://www.cs.washington.edu/research/projects/cecil/www/pubs/
\paragraph{Multimethods} 
\cite{MS02} and collaborators present a sequence of
systems~\cite{Chambers:1992:OMC,Chambers:1994:TMM,MS02} with statically-typed multimethods
and modular type checking.  In contrast to Typed Clojure, in these
system methods declare the types of arguments that they expect which
corresponds to exclusively using \clj{class} as the dispatch function
in Typed Clojure. However, Typed Clojure does not attempt to rule out
failed dispatches at runtime.
